ANGSD-wrapper: utilities for analyzing next generation sequencing data

### 2016-01-29 manuscript.tex
addressed comments

\begin{abstract}
High throughput sequencing has changed many aspects of population genetics, molecular ecology, and related fields, affecting both experimental design and data analysis.
The software package ANGSD allows users to perform a number of population genetic analyses on high-throughput sequencing data. 
The package is specifically designed to produce more accurate results for samples with low sequencing depth and makes use of full genome data while handling a wide array of sampling and experimental designs.
Here we present ANGSD-wrapper, a user-friendly interface for running ANGSD and visualizing results.
ANGSD-wrapper includes a number of 'wrapper' scripts that facilitate configuration and execution of multi-step analyses. ANGSD-wrapper also provides interactive graphing of ANGSD results to enhance data exploration.
We demonstrate the usefulness of ANGSD-wrapper by analyzing resequencing data from populations of wild and domesticated \textit{Zea}. 
ANGSD-wrapper is freely available from \url{https://github.com/mojaveazure/angsd-wrapper}.
\end{abstract}

@@

Here we present ANGSD-wrapper, a user-friendly interface to stable versions of ANGSD. 
ANGSD-wrapper takes the form of a set of configuration files and `wrapper' scripts (Figure \ref{fig:supp2}) that streamline the execution of multi-step pipelines required for data analysis in ANGSD. 
ANGSD-wrapper also eases the configuration of ANGSD-related programs such as \tjyk{Fumagali citation doesn't show up properly}\ph{I think the citation is just 'Fumagali' and the link to the webpage. If so, then it works for me -PH}ngsPopGen \citep{ngsPopGen}, ngsF\citep{vieira2013estimating}, and ngsAdmix \citep{pmid24026093}. 
\ph{This could be reworded} Furthermore, ANGSD-wrapper allows for easy population genetic analysis across the sequence data provided, or hone in on a certain region or subset of regions, allowing a highly-tailored experience.
Additionally, the wrapper scripts are written against `frozen' versions of ANGSD and supporting tools, for consistency of analysis. 
Because the large volume of data associated with high throughput sequence analysis is often difficult to visualize, ANGSD-wrapper also provides a suite of interactive visualization tools to plot results and explore patterns at multiple scales.  
We demonstrate some of the analyses possible using ANGSD-wrapper when applied to low-coverage whole-genome data from domesticated maize and two related wild teosinte subspecies. 
ANGSD-wrapper is freely available from \url{https://github.com/mojaveazure/angsd-wrapper}.

\section*{Methods}
ANGSD-wrapper is a set of configuration files and scripts written primarily in the Bash scripting language. 
The scripts can be run either on a standalone computer with a UNIX terminal, \ph{has anyone tested AW on a cluster?} or on computing clusters where they can be submitted to a queuing system such as SGE \citep{Microsystems):2001:SGE:560889.792378}, Slurm \citep{Jette02slurm:simple} or TORQUE \citep{Staples:2006:TRM:1188455.1188464}.  
A Python installation \citep{Python} is required for some light, dynamic pre-processing of the data.
An installation of the statistical software R \citep{Rcitation} is required to make use of the visualization tools incorporated in ANGSD-wrapper.  
The visualization portion of ANGSD-wrapper also requires installation of the R packages \texttt{Shiny} \citep{shiny}, \texttt{APE} \citep{APE}, \texttt{Lattice} \citep{lattice}, \texttt{Hmisc} \citep{Hmisc}, \texttt{data.table} \citep{datatable}, \texttt{DT} \citep{javascriptDT}, and \texttt{shinythemes} \citep{shinythemes}, as well as \texttt{genomeIntervals} \citep{genomeIntervals} from Bioconductor \citep{bioconductor} for the visualization package.
These dependent packages are installed automatically upon first run of visualization interface.

@@

\ph{Necessary?}The visualization software included with ANGSD-wrapper is contained within it's own directory called ``shinyGraphing.''
This application is started from within ANGSD-wrapper and launched locally from a standard web browser. %This application must be started in R and can be accessed locally from a web browser. --Nope
This software provides a graphical user interface (GUI) to quickly and interactively plot results obtained from ANGSD-wrapper.  
Each tab in the GUI contains plots for different ANGSD methods.

@@

\begin{figure}
\centering
\includegraphics[width=0.8\linewidth]{figures/figure3big.pdf}
\caption{Summary statistics for {\it Zea mays}. Derived site frequency spectra for A. maize and B. teosinte. C. distribution of pairwise differences for maize (blue) and teosinte (orange). Pairwise diversity results are visualized separately from the interactive graphics and colors were added to A and B using a custom script.}
\label{fig:figure3}
\end{figure}
\tjyk{Should this start a "Results" section?}
As a demonstration of analyses in ANGSD-wrapper, we explore patterns of diversity in a single genomic region of domesticated maize and wild teosinte. 
We used a subset of the resequenced samples from the Maize HapMap2 project (Table \ref{tab:samples}) and calculated summary statistics using a 10Mb region on chromosome 10 \citep{chia2012maize}. 
The data are available at \url{https://figshare.com/articles/Example_Data_tar_bz2/2063442}. 
In the following we refer to methods listed in Table \ref{tab:methods} when describing analyses.

We first use the ``SFS'' method to estimate the site frequency spectrum (SFS) of both maize and its wild progenitor \textit{Zea mays} ssp. \textit{parviglumis}, assuming an inbreeding coefficient of $F=1$ for these highly inbred samples. 
The SFS and diversity statistics were calculated using ancestral states inferred in the ``Ancestral Sequence'' method from a single resequenced genome of \textit{Tripsacum dactyloides}.  
We show that the maize SFS is skewed towards more intermediate-frequency variants (Figure \ref{fig:figure3}A-B, Tajima's D of 0.2 and 0.0085 in maize and teosinte, respectively), likely a result of the bottleneck associated with maize domestication \citep{eyre1998investigation, Beissinger031666}.
Using the ``Thetas'' method we find further evidence of the domestication bottleneck, with mean levels of pairwise nucleotide diversity in this region in maize $\approx 25\%$ lower than in teosinte (0.0061 and 0.0082, respectively; Figure \ref{fig:figure3}C). 
Using the ``2DSFS method,'' which includes an \fst calculation, we find a mean \fst in this region of 0.116, nearly identical to the  genome-wide value of 0.11 reported in \cite{hufford2012comparative}.   
There are no genes in this region that have  been identified as potential domestication candidates, consistent with the lack of extended regions of high \fst in our analysis (Figure \ref{fig:suppfst}).

\begin{figure}
\centering
\includegraphics[width=0.8\linewidth]{figures/mt4labeled.pdf}
\caption{Admixture analysis for {\it Zea mays} ssp. {\it mays}, {\it Zea mays} ssp. {\it parviglumis}, and {\it Zea mays} ssp. {\it mexicana} (\textit{mex}) with colors representing the K=4 source populations. 
Individuals are shown in the same order as Table \ref{tab:samples}. }
\label{fig:admixture}
\end{figure}
\jri{what are vertical lines in pi density plot? need to be consistent about saying "nucleotide diversity" vs "pairwise diversity" vs "pairwise differences" etc.  I vote for "pairwise nucleotide diversity"}
Finally, we include two samples of the related wild teosinte \textit{Zea mays} ssp. \textit{mexicana} to assess evidence for admixture.  
Using the ``Admixture'' method, which implements an estimate of admixture proportion from genotype likelihoods \citep{pmid24026093}, we identify structure within domesticated maize separating three high-latitude temperate landraces from the other tropical accessions (Figure \ref{fig:admixture}). 
This analysis is a supervised method that assigns proportions of the each individual's genome to a putative source population, represented by the colors. 
We ran this analysis over a range of number of ancestral populations (K= 2, 3, 4; See Figure \ref{fig:suppadmix}) and chose the value with the maximum likelihood while keeping the number of parameters as low as possible. 
In maize, we see that most individuals come from one source population (dark grey or dark blue), which is evidence for no admixture between these lowland maize samples and  ssp. \textit{mexicana}, and is consistent with an independent analysis from SNP genotyping \citep{hufford2013genomic}.
\textit{Zea mays} ssp. \textit{mexicana} clusters into its own group (orange), along with a single accession of ssp. \textit{parviglumis} collected from a region in which many teosinte populations appear to be the result of admixture between the two subspecies \citep{fang2012megabase}.  
A single ssp. \textit{parviglumis} accession from the Northern extent of the range does not appear to be well-classified with these data, likely due to our relatively limited geographic and genomic sampling.

>>>>

\begin{abstract}
High throughput sequencing has changed many aspects of population genetics, molecular ecology, and related fields, affecting both experimental design and data analysis.
The software package ANGSD allows users to perform a number of population genetic analyses on high-throughput sequencing data.
% \cl{added text}
ANGSD uses probabilistic approaches to calculate genome wide descriptive statistics.
The package is specifically designed to produce more accurate results for samples with low sequencing depth and makes use of full genome data while handling a wide array of sampling and experimental designs.
Here we present ANGSD-wrapper, a user-friendly interface for running ANGSD and visualizing results.
% \cl{added text}
This software supports multiple types of analyses such as Site Frequency Spectrum, Genotype Likelihood estimations, Thetas diversity statistics and neutrality tests, Principal Component Analysis, Admixture, and Inbreeding coefficient calculations.
ANGSD-wrapper includes a number of 'wrapper' scripts that facilitate configuration and execution of multi-step analyses. ANGSD-wrapper also provides interactive graphing of ANGSD results to enhance data exploration.
We demonstrate the usefulness of ANGSD-wrapper by analyzing resequencing data from populations of wild and domesticated \textit{Zea}. 
ANGSD-wrapper is freely available from \url{https://github.com/mojaveazure/angsd-wrapper}.
\end{abstract}

@@

Here we present ANGSD-wrapper, a user-friendly interface to stable versions of ANGSD. 
ANGSD-wrapper takes the form of a set of configuration files and `wrapper' scripts (Figure \ref{fig:supp2}) that streamline the execution of multi-step pipelines required for data analysis in ANGSD. 
ANGSD-wrapper also eases the configuration of ANGSD-related programs such as ngsPopGen \citep{ngsPopGen}, ngsF\citep{vieira2013estimating}, and ngsAdmix \citep{pmid24026093}. %\tjyk{Fumagali citation doesn't show up properly}\ph{I think the citation is just 'Fumagali' and the link to the webpage. If so, then it works for me -PH} \ad{yes, there's no citation other than the github repo (and no DOI there)}
% \ph{This could be reworded} 
Furthermore, ANGSD-wrapper allows for easy population genetic analysis across the sequence data provided, or hone in on a certain region or subset of regions, allowing a highly-tailored experience.
Additionally, the wrapper scripts are written against `frozen' versions of ANGSD and supporting tools, for consistency of analysis. 
Because the large volume of data associated with high throughput sequence analysis is often difficult to visualize, ANGSD-wrapper also provides a suite of interactive visualization tools to plot results and explore patterns at multiple scales.  
We demonstrate some of the analyses possible using ANGSD-wrapper when applied to low-coverage whole-genome data from domesticated maize and two related wild teosinte subspecies. 
ANGSD-wrapper is freely available from \url{https://github.com/mojaveazure/angsd-wrapper}.

\section*{Methods}
ANGSD-wrapper is a set of configuration files and scripts written primarily in the Bash scripting language. 
The scripts can be run either on a standalone computer with a UNIX terminal, or on computing clusters where they can be submitted to a queuing system such as SGE \citep{Microsystems):2001:SGE:560889.792378}, Slurm \citep{Jette02slurm:simple} or TORQUE \citep{Staples:2006:TRM:1188455.1188464}.  % \ph{has anyone tested AW on a cluster?} \ad{yep I have. I logged into a compute node via srun and ran it that way}
A Python installation \citep{Python} is required for some light, dynamic pre-processing of the data.
An installation of the statistical software R \citep{Rcitation} is required to make use of the visualization tools incorporated in ANGSD-wrapper.  
The visualization portion of ANGSD-wrapper also requires installation of the R packages \texttt{Shiny} \citep{shiny}, \texttt{APE} \citep{APE}, \texttt{Lattice} \citep{lattice}, \texttt{Hmisc} \citep{Hmisc}, \texttt{data.table} \citep{datatable}, \texttt{DT} \citep{javascriptDT}, and \texttt{shinythemes} \citep{shinythemes}, as well as \texttt{genomeIntervals} \citep{genomeIntervals} from Bioconductor \citep{bioconductor} for the visualization package.
These dependent packages are installed automatically upon first run of visualization interface.

@@

% \ph{Necessary?} \ad{changed the wording}
The visualization software included with ANGSD-wrapper is contained in a directory called ``shinyGraphing.''
This application is started from within ANGSD-wrapper and launched locally from a standard web browser. 
This software provides a graphical user interface (GUI) to quickly and interactively plot results obtained from ANGSD-wrapper.  
Each tab in the GUI contains plots for different ANGSD methods.

@@

\section*{Results}

\begin{figure}
\centering
\includegraphics[width=0.8\linewidth]{figures/figure3big.pdf}
\caption{Summary statistics for {\it Zea mays}. Derived site frequency spectra for A. maize and B. teosinte. C. distribution of pairwise nucleotide diversity for maize (blue) and teosinte (orange). Pairwise nucleotide diversity results are visualized separately from the interactive graphics and colors were added to A and B using a custom script.}
\label{fig:figure3}
\end{figure}

%\tjyk{Should this start a "Results" section?} \ad{yes I think so.}

As a demonstration of analyses in ANGSD-wrapper, we explore patterns of pairwise nucleotide diversity in a single genomic region of domesticated maize and wild teosinte. 
We used a subset of the resequenced samples from the Maize HapMap2 project (Table \ref{tab:samples}) and calculated summary statistics using a 10Mb region on chromosome 10 \citep{chia2012maize}. 
The data are available at \url{https://figshare.com/articles/Example_Data_tar_bz2/2063442}. 
In the following we refer to methods listed in Table \ref{tab:methods} when describing analyses.

We first use the ``SFS'' method to estimate the site frequency spectrum (SFS) of both maize and its wild progenitor \textit{Zea mays} ssp. \textit{parviglumis}, assuming an inbreeding coefficient of $F=1$ for these highly inbred samples. 
The SFS and diversity statistics were calculated using ancestral states inferred in the ``Ancestral Sequence'' method from a single resequenced genome of \textit{Tripsacum dactyloides}.  
We show that the maize SFS is skewed towards more intermediate-frequency variants (Figure \ref{fig:figure3}A-B, Tajima's D of 0.2 and 0.0085 in maize and teosinte, respectively), likely a result of the bottleneck associated with maize domestication \citep{eyre1998investigation, Beissinger031666}.
Using the ``Thetas'' method we find further evidence of the domestication bottleneck, with mean levels of pairwise nucleotide diversity in this region in maize $\approx 25\%$ lower than in teosinte (0.0061 and 0.0082, respectively). Figure \ref{fig:figure3}C shows the distribution of pairwise nucleotide diversity in each window within our region of interest and each vertical line represents the number of windows with the corresponding value of $\pi$. 
Using the ``2DSFS method,'' which includes an \fst calculation, we find a mean \fst in this region of 0.116, nearly identical to the  genome-wide value of 0.11 reported in \cite{hufford2012comparative}.   
There are no genes in this region that have  been identified as potential domestication candidates, consistent with the lack of extended regions of high \fst in our analysis (Figure \ref{fig:suppfst}).

\begin{figure}
\centering
\includegraphics[width=0.8\linewidth]{figures/mt4labeled.pdf}
\caption{Admixture analysis for {\it Zea mays} ssp. {\it mays}, {\it Zea mays} ssp. {\it parviglumis}, and {\it Zea mays} ssp. {\it mexicana} (\textit{mex}) with colors representing the K=4 source populations. 
Individuals are shown in the same order as Table \ref{tab:samples}. }
\label{fig:admixture}
\end{figure}
% \jri{what are vertical lines in pi density plot? need to be consistent about saying "nucleotide diversity" vs "pairwise diversity" vs "pairwise differences" etc.  I vote for "pairwise nucleotide diversity"} \ad{changed.}
Finally, we include two samples of the related wild teosinte \textit{Zea mays} ssp. \textit{mexicana} to assess evidence for admixture.  
Using the ``Admixture'' method, which implements an estimate of admixture proportion from genotype likelihoods \citep{pmid24026093}, we identify structure within domesticated maize separating three high-latitude temperate landraces from the other tropical accessions (Figure \ref{fig:admixture}). 
This analysis is a supervised method that assigns proportions of the each individual's genome to a putative source population, represented by the colors. 
We ran this analysis over a range of number of ancestral populations (K= 2, 3, 4; See Figure \ref{fig:suppadmix}) and chose the value with the maximum likelihood while keeping the number of parameters as low as possible. 
In maize, we see that most individuals come from one source population (dark grey or dark blue), which is evidence for no admixture between these lowland maize samples and  ssp. \textit{mexicana}, and is consistent with an independent analysis from SNP genotyping \citep{hufford2013genomic}.
\textit{Zea mays} ssp. \textit{mexicana} clusters into its own group (orange), along with a single accession of ssp. \textit{parviglumis} collected from a region in which many teosinte populations appear to be the result of admixture between the two subspecies \citep{fang2012megabase}.  
A single ssp. \textit{parviglumis} accession from the Northern extent of the range does not appear to be well-classified with these data, likely due to our relatively limited geographic and genomic sampling.
